rbeast2: BEAUti 2, BEAST2 and Tracer for R

### 2018-02-15 article.tex
Renamed package to 'rbeast2' from 'babette'

\title{babette: BEAUti 2, BEAST2 and Tracer for R}
\author{Rich\`el J.C. Bilderbeek, Rampal S. Etienne}

@@

  \textbf{1. }
    In the field of phylogenetics, 
     BEAST2 is one of the most widely used software tools. 
     It comes with the graphical programs BEAUti 2 to
     facilitate the creation of BEAST2 configuration files, and
     with DensiTree and Tracer, to interpret BEAST2's output files. 
     However, when many different alignments or model 
     setups are required, a workflow of GUI programs is cumbersome. \\
  \textbf{2. }
    Here, we present a free, libre and open-source package, \verb;babette;, 
    'BEAUti 2, BEAST2 and Tracer for R', for the R programming language. 
    \verb;babette; creates BEAST2 input files, runs BEAST2 and parses its results, 
    all from an R function call. \\
  \textbf{3. }
    We describe \verb;babette;'s usage, the novel functionality it provides
    compared to the tools it is alternative of, and give some examples. \\
  \textbf{4. }
    As \verb;babette; is designed to be of high quality and extendable, 
    we conclude by describing the further development of the package \\
\end{abstract}

@@

Here, to provide such functionality we present \verb;babette;, 
’BEAUti 2, BEAST2 and Tracer for R’, 
which creates BEAST2 configuration files,
runs BEAST2, and analyzes its results,
all from an R function call. This
will save time, tedious mouse clicking and 
reduces the chances of errors in such repetitive actions.
The interface of \verb;babette; mimics the tools it
is inspired on. This
familiarity helps both beginner and experienced BEAST2 users 
to make the step from those tools to \verb;babette;.
\verb;babette; enables the creation of a single-script 
pipeline from sequence alignments to posterior analysis in R.

\verb;babette; is the first R package that unifies the
full workflow of working with BEAST2. An
example to create a BEAST2 input file is 
\verb;BEASTmasteR; \cite{beastmaster}.
Also \verb;BEASTmasteR; allows 
to create BEAST2 configuration files from R. 
The difference is that \verb;babette; 
has its focus on DNA alignments and ultrametric trees, 
where \verb;BEASTmasteR; is used for 
morphological traits and tip-dating.
Examples of R packages to parse the BEAST2 output files
are \verb;rBEAST; \cite{rBEAST} and 
\verb;RBEAST; \cite{RBEAST}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  \centering
  \resizebox {0.8\columnwidth} {!} {
    \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=6cm, semithick]   
    \tikzstyle{every state}=[]
    \node[state] (A) [rectangle] {
      \includegraphics[width=0.2\textwidth]{alignment.png}
    };   
    \node[state] (AL) [above left=-0.25cm and -0.25cm of A, fill=white] {
      1
    };   
    \node[state] (B) [right of = A, rectangle] {
      \includegraphics[height=0.15\textheight]{BeautiSiteModel.png}
    };   
    \node[state] (BL) [above left=-0.25cm and -0.25cm of B, fill=white] {
      2
    };   
    \node[state] (E) [below of=A, rectangle] {
      \includegraphics[height=0.1\textheight]{file2.png}
    };   
    \node[state] (EL) [above left=-0.25cm and -0.25cm of E, fill=white] {
      3
    };   
    \node[state] (C) [right of = E, rectangle] {
      \includegraphics[width=0.3\textwidth]{beast_logo.png}
    };   
    \node[state] (CL) [above left=-0.25cm and -0.25cm of C, fill=white] {
      4
    };   
    \node[state] (F) [rectangle, below of=E] {
      \includegraphics[height=0.1\textheight]{file.png}
    };
    \node[state] (FL) [above left=-0.25cm and -0.25cm of F, fill=white] {
      5
    };   
    \node[state] (H) [rectangle, below of=F] {
      \includegraphics[width=0.3\textwidth]{DensiTreeExample2.jpg}
    };
    \node[state] (HL) [above left=-0.25cm and -0.25cm of H, fill=white] {
      6
    };   
    \node[state] (I) [rectangle, right of=H] {
      \includegraphics[width=0.6\textwidth]{tracer_example_output.png}
    };
    \node[state] (IL) [above left=-0.25cm and -0.25cm of I, fill=white] {
      7
    };   
    \path 
      (A) edge [anchor = west] node {} (E)
      (E) edge [anchor = west] node {} (F)
      (F) edge [anchor = west] node {} (H)
      (F) edge [anchor = west] node {} (I)
    ; 
    \path [-o,draw] (B) -- ($ (A) !.5! (E) $);
    \path [-o,draw] (C) -- ($ (E) !.5! (F) $);
    \end{tikzpicture}
  }
  \caption{
    Workflow using GUI tools. From an alignment (1) and BEAUti (2), 
    a BEAST2 configuration file (3) is created. BEAST2 (4) uses that file
    to infer a posterior, storing it in multiple files (5). These results
    are visualized using DensiTree (6) and Tracer (7). babette allows
    for the same workflow, all from an R function call.
  }
  \label{fig:workflow}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\verb;babette; is written in the R programming language \cite{R}
and enables the full BEAST2 workflow from an R function call,
in a similar way to what BEAUti, DensiTree and Tracer do.

\verb;babette;'s main function is \verb;run_beast2;, which, 
configures BEAST2, runs it and parses its output. 
\verb;run_beast2; needs at least the name of a 
FASTA file containing a DNA alignment. 
The default settings for the other arguments of \verb;run_beast2; 
are identical to BEAUti's and BEAST2's default settings.
Per alignment, a site model, clock model and tree prior can be chosen.
Multiple alignments can be used, each with its own (unlinked) site model, 
clock model and tree prior.

\verb;babette; currently has 61 exported functions to set up  
a BEAST2 configuration file. 
\verb;babette; is an alternative for a majority of BEAUti use cases, 
but does not yet support the full functionality of BEAUti. 
Because of BEAUti's high number of plugins, 
\verb;babette; uses a software architecture that expects to be extended.

\verb;babette; has 7 exported function to run and help run BEAST2.
One function is used to run BEAST2, others
allow the user to check if a BEAST2 configuration file is indeed valid.

\verb;babette; has 20 exported function to parse the BEAST2 output
files and analyse the created posterior. \verb;babette; gives the
same ESSes and summary statistics as Tracer. The data is formatted
as such, that it can easily be visualized using \verb;ggplot2; (for a trace,
similar to Tracer)
or \verb;phangorn; \cite{phangorn} (for the phylogenies in a posterior, similar to DensiTree).

@@

\begin{lstlisting}[language=R, floatplacement=H]
library(babette)
\end{lstlisting}

BEAUti, and likewise \verb;babette;, needs at least a FASTA filename
to produce a BEAST2 configuration file. 
In BEAUti, this is achieved by loading a FASTA file, 
then saving an output file using a common
save file dialog. After this, BEAST2 needs to be invoked
on the created configuration file to create multiple files
storing the created posterior. Finally, these output
files must be parsed by either Tracer of DensiTree.

In \verb;babette;, the same is achieved by:

@@

This code will create a (temporary) BEAST2 configuration file,
using a FASTA file with name \verb;alignment.fas;, 
using the same default settings as BEAUti.
\verb;babette; will then execute BEAST2 with that file, and
parses the output. The output, named \verb;out;, is a list of
parameter estimates, posterior phylogenies (one for each alignment)
and operator acceptances.

The default settings for BEAUti (and thus \verb;babette;) are, 
among others, to use a Jukes-Cantor site model \cite{cantor1969mammalian}, 
a strict clock, and a Yule birth tree prior \cite{yule}.

@@

\begin{table}[]
\centering
\begin{tabular}{ | l | l | }
\hline
\textbf{Name} & \textbf{Description} \\
\hline
\verb;create_beast2_input_file; & Creates a BEAST2 input file \\
\hline
\verb;create_gtr_site_model; & Create a GTR site model \\
\verb;create_hky_site_model; & Create an HKY site model \\
\verb;create_jc69_site_model; & Create a Jukes-Cantor site model \\
\verb;create_tn93_site_model; & Create a TN93 site model \\
\hline
\verb;create_rln_clock_model; & Create a relaxed log-normal clock model \\
\verb;create_strict_clock_model; & Create a strict clock model \\
\hline
\verb;create_bd_tree_prior; & Create a birth-death tree prior \\
\verb;create_cbs_tree_prior; & Create a coalescent Bayesian skyline tree prior \\
\verb;create_ccp_tree_prior; & Create a coalescent constant-population tree prior \\
\verb;create_cep_tree_prior; & Create a coalescent exponential-population tree prior \\
\verb;create_yule_tree_prior; & Create a Yule tree prior \\
\hline
\verb;create_beta_distr; & Create a beta distribution \\
\verb;create_exp_distr; & Create an exponential distribution \\
\verb;create_gamma_distr; & Create a gamma distribution \\
\verb;create_inv_gamma_distr; & Create an inverse gamma distribution \\
\verb;create_laplace_distr; & Create a Laplace distribution \\
\verb;create_log_normal_distr; & Create a log-normal distribution \\
\verb;create_normal_distr; & Create a normal distribution \\
\verb;create_one_div_x_distr; & Create a 1/X distribution \\
\verb;create_poisson_distr; & Create a Poisson distribution \\
\verb;create_uniform_distr; & Create a uniform distribution \\
\hline
\end{tabular}
\caption{babette's main functions}
\label{tab:functions}
\end{table}

@@

\verb;babette; also uses the same default prior distributions as BEAUti 
for each of the site models, clock models and tree priors. 
For example, by default, a Yule tree prior assumes the birth rate 
follows a uniform distribution, 
from minus infinity to plus infinity. 
This assumption implies that negative and positive birth rates are just as likely, 
where a negative birth rate is biologically impossible (note that 
in practice, this usually works out just fine).
One may prefer an exponential distribution instead, 
as this would assume only positive birth rates, 
and makes high birth rates unlikely.

The following script shows how to do this in \verb;babette;:

@@

Our initial motivation to create \verb;babette; 
is that we wanted to fix the crown age of a phylogeny.
BEAUti assumes that a phylogeny has a crown age that needs to be jointly-estimated
with the phylogeny and other parameters. It does not allow for fixing
the crown age. Without \verb;babette;, one needs to manually edit the BEAST2 
configuration file, which is tedious and prone to errors. 
Fixing the crown ages is especially useful for theoretical experiments,
as this allows for one less source of variation.

This is how to specify a fixed crown age with \verb;babette;:

@@

\verb;babette; allows for the same functionality of Tracer. 
Tracer, among others, shows the effective sample
sizes of each posterior's variables. As an MCMC run needs to
converge first, Tracer discards the first 10\% of all parameter
estimates. This is called the "burn-in". To remove the burn-in
in \verb;babette;:

@@

Then, to calculate the effective sample sizes in \verb;babette;:

@@

Tracer displays multiple summary statistics for each
estimated variable: the mean and its standard error, standard deviation,
variance, median, mode, geometric mean, 95\% highest posterior density interval, 
auto-correlation time and effective sample size. To obtain all these summary statistics 
of, for example, the estimated birth rate in \verb;babette;:

@@

\verb;babette; allows for the same functionality of DensiTree.
DensiTree displays the phylogenies in a posterior at the same
time scale over one another, allowing to see the uncertainty in
topology and branch lengths. To visualize the same in \verb;babette;:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{babette development and other resources}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\verb;babette; is free, libre and open source software available from the official R package archive at 
\url{http://cran.r-project.org/src/contrib/PACKAGES.html\#babette}
and is licensed under the GNU General Public License v3.0.

\verb;babette; uses the Travis CI \cite{travis} 
continuous integration service, which is known to significantly 
increase the the number of bugs exposed \cite{vasilescu2015} and increases
the speed at which new features are added \cite{vasilescu2015}.
\verb;babette; has a 100\% code coverage, which correlates with code quality \cite{horgan1994,del1995correlation}. 
\verb;babette; follows Hadley Wickham's style guide \cite{style_guide}, which improves software quality \cite{fang2001}.
\verb;babette; is dependent on multiple packages, which are 
\verb;APE; \cite{APE}, 
\verb;beautier; \cite{beautier},
\verb;beastier; \cite{beastier},
\verb;devtools; \cite{devtools},
\verb;geiger; \cite{GEIGER},
\verb;ggplot2; \cite{ggplot2},
\verb;knitr; \cite{knitr},
\verb;phangorn; \cite{phangorn},
\verb;rmarkdown; \cite{rmarkdown},
\verb;seqinr; \cite{seqinr},
\verb;stringr; \cite{stringr},
\verb;testit; \cite{testit} and 
\verb;tracerer; \cite{tracerer} and
\verb;TreeSim; \cite{TreeSim}.

\verb;babette;'s development takes place on GitHub \cite{github},
\url{https://github.com/richelbilderbeek/babette}, 
which accommodates collaboration \cite{perez2016ten} 
and improves transparency \cite{gorgolewski2016practical}.
\verb;babette;'s GitHub facilitates feature requests and has guidelines how to do so.

\verb;babette;'s documentation is extensive. All functions are documented
in the package's internal documentation. For quick use, 
each exported function shows a minimal example. 
For easy exploration, each exported function's documentation links to related functions.
Additionally, \verb;babette; has a vignette that demonstrates extensively how
to use it. The GitHub documentation helps to get started, with a dozen examples 
of BEAUti screenshots with equivalent \verb;babette; code.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Citation of babette}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Scientists using \verb;babette; in a published paper can cite this
article, and/or cite the \verb;babette; package 
directly. To obtain this citation from within an R script, use:

\begin{lstlisting}[language=R]
> citation("babette")
\end{lstlisting}

>>>>

\title{rbeast2: BEAUti 2, BEAST2 and Tracer for R}
\author{Rich\`el J.C. Bilderbeek, Rampal S. Etienne}

@@

  \textbf{1. }
    In the field of phylogenetics, 
     BEAST2 is one of the most widely used software tools. 
     It comes with the graphical programs BEAUti 2 to
     facilitate the creation of BEAST2 configuration files, and
     with DensiTree and Tracer, to interpret BEAST2's output files. 
     However, when many different alignments or model 
     setups are required, a workflow of GUI programs is cumbersome. \\
  \textbf{2. }
    Here, we present a free, libre and open-source package, \verb;rbeast2;, 
    'BEAUti 2, BEAST2 and Tracer for R', for the R programming language. 
    \verb;rbeast2; creates BEAST2 input files, runs BEAST2 and parses its results, 
    all from an R function call. \\
  \textbf{3. }
    We describe \verb;rbeast2;'s usage, the novel functionality it provides
    compared to the tools it is alternative of, and give some examples. \\
  \textbf{4. }
    As \verb;rbeast2; is designed to be of high quality and extendable, 
    we conclude by describing the further development of the package \\
\end{abstract}

@@

Here, to provide such functionality we present \verb;rbeast2;, 
’BEAUti 2, BEAST2 and Tracer for R’, 
which creates BEAST2 configuration files,
runs BEAST2, and analyzes its results,
all from an R function call. This
will save time, tedious mouse clicking and 
reduces the chances of errors in such repetitive actions.
The interface of \verb;rbeast2; mimics the tools it
is inspired on. This
familiarity helps both beginner and experienced BEAST2 users 
to make the step from those tools to \verb;rbeast2;.
\verb;rbeast2; enables the creation of a single-script 
pipeline from sequence alignments to posterior analysis in R.

\verb;rbeast2; is the first R package that unifies the
full workflow of working with BEAST2. An
example to create a BEAST2 input file is 
\verb;BEASTmasteR; \cite{beastmaster}.
Also \verb;BEASTmasteR; allows 
to create BEAST2 configuration files from R. 
The difference is that \verb;rbeast2; 
has its focus on DNA alignments and ultrametric trees, 
where \verb;BEASTmasteR; is used for 
morphological traits and tip-dating.
Examples of R packages to parse the BEAST2 output files
are \verb;rBEAST; \cite{rBEAST} and 
\verb;RBEAST; \cite{RBEAST}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
  \centering
  \resizebox {0.8\columnwidth} {!} {
    \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=6cm, semithick]   
    \tikzstyle{every state}=[]
    \node[state] (A) [rectangle] {
      \includegraphics[width=0.2\textwidth]{alignment.png}
    };   
    \node[state] (AL) [above left=-0.25cm and -0.25cm of A, fill=white] {
      1
    };   
    \node[state] (B) [right of = A, rectangle] {
      \includegraphics[height=0.15\textheight]{BeautiSiteModel.png}
    };   
    \node[state] (BL) [above left=-0.25cm and -0.25cm of B, fill=white] {
      2
    };   
    \node[state] (E) [below of=A, rectangle] {
      \includegraphics[height=0.1\textheight]{file2.png}
    };   
    \node[state] (EL) [above left=-0.25cm and -0.25cm of E, fill=white] {
      3
    };   
    \node[state] (C) [right of = E, rectangle] {
      \includegraphics[width=0.3\textwidth]{beast_logo.png}
    };   
    \node[state] (CL) [above left=-0.25cm and -0.25cm of C, fill=white] {
      4
    };   
    \node[state] (F) [rectangle, below of=E] {
      \includegraphics[height=0.1\textheight]{file.png}
    };
    \node[state] (FL) [above left=-0.25cm and -0.25cm of F, fill=white] {
      5
    };   
    \node[state] (H) [rectangle, below of=F] {
      \includegraphics[width=0.3\textwidth]{DensiTreeExample2.jpg}
    };
    \node[state] (HL) [above left=-0.25cm and -0.25cm of H, fill=white] {
      6
    };   
    \node[state] (I) [rectangle, right of=H] {
      \includegraphics[width=0.6\textwidth]{tracer_example_output.png}
    };
    \node[state] (IL) [above left=-0.25cm and -0.25cm of I, fill=white] {
      7
    };   
    \path 
      (A) edge [anchor = west] node {} (E)
      (E) edge [anchor = west] node {} (F)
      (F) edge [anchor = west] node {} (H)
      (F) edge [anchor = west] node {} (I)
    ; 
    \path [-o,draw] (B) -- ($ (A) !.5! (E) $);
    \path [-o,draw] (C) -- ($ (E) !.5! (F) $);
    \end{tikzpicture}
  }
  \caption{
    Workflow using GUI tools. From an alignment (1) and BEAUti (2), 
    a BEAST2 configuration file (3) is created. BEAST2 (4) uses that file
    to infer a posterior, storing it in multiple files (5). These results
    are visualized using DensiTree (6) and Tracer (7). rbeast2 allows
    for the same workflow, all from an R function call.
  }
  \label{fig:workflow}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\verb;rbeast2; is written in the R programming language \cite{R}
and enables the full BEAST2 workflow from an R function call,
in a similar way to what BEAUti, DensiTree and Tracer do.

\verb;rbeast2;'s main function is \verb;run_beast2;, which, 
configures BEAST2, runs it and parses its output. 
\verb;run_beast2; needs at least the name of a 
FASTA file containing a DNA alignment. 
The default settings for the other arguments of \verb;run_beast2; 
are identical to BEAUti's and BEAST2's default settings.
Per alignment, a site model, clock model and tree prior can be chosen.
Multiple alignments can be used, each with its own (unlinked) site model, 
clock model and tree prior.

\verb;rbeast2; currently has 61 exported functions to set up  
a BEAST2 configuration file. 
\verb;rbeast2; is an alternative for a majority of BEAUti use cases, 
but does not yet support the full functionality of BEAUti. 
Because of BEAUti's high number of plugins, 
\verb;rbeast2; uses a software architecture that expects to be extended.

\verb;rbeast2; has 7 exported function to run and help run BEAST2.
One function is used to run BEAST2, others
allow the user to check if a BEAST2 configuration file is indeed valid.

\verb;rbeast2; has 20 exported function to parse the BEAST2 output
files and analyse the created posterior. \verb;rbeast2; gives the
same ESSes and summary statistics as Tracer. The data is formatted
as such, that it can easily be visualized using \verb;ggplot2; (for a trace,
similar to Tracer)
or \verb;phangorn; \cite{phangorn} (for the phylogenies in a posterior, similar to DensiTree).

@@

\begin{lstlisting}[language=R, floatplacement=H]
library(rbeast2)
\end{lstlisting}

BEAUti, and likewise \verb;rbeast2;, needs at least a FASTA filename
to produce a BEAST2 configuration file. 
In BEAUti, this is achieved by loading a FASTA file, 
then saving an output file using a common
save file dialog. After this, BEAST2 needs to be invoked
on the created configuration file to create multiple files
storing the created posterior. Finally, these output
files must be parsed by either Tracer of DensiTree.

In \verb;rbeast2;, the same is achieved by:

@@

This code will create a (temporary) BEAST2 configuration file,
using a FASTA file with name \verb;alignment.fas;, 
using the same default settings as BEAUti.
\verb;rbeast2; will then execute BEAST2 with that file, and
parses the output. The output, named \verb;out;, is a list of
parameter estimates, posterior phylogenies (one for each alignment)
and operator acceptances.

The default settings for BEAUti (and thus \verb;rbeast2;) are, 
among others, to use a Jukes-Cantor site model \cite{cantor1969mammalian}, 
a strict clock, and a Yule birth tree prior \cite{yule}.

@@

\begin{table}[]
\centering
\begin{tabular}{ | l | l | }
\hline
\textbf{Name} & \textbf{Description} \\
\hline
\verb;create_beast2_input_file; & Creates a BEAST2 input file \\
\hline
\verb;create_gtr_site_model; & Create a GTR site model \\
\verb;create_hky_site_model; & Create an HKY site model \\
\verb;create_jc69_site_model; & Create a Jukes-Cantor site model \\
\verb;create_tn93_site_model; & Create a TN93 site model \\
\hline
\verb;create_rln_clock_model; & Create a relaxed log-normal clock model \\
\verb;create_strict_clock_model; & Create a strict clock model \\
\hline
\verb;create_bd_tree_prior; & Create a birth-death tree prior \\
\verb;create_cbs_tree_prior; & Create a coalescent Bayesian skyline tree prior \\
\verb;create_ccp_tree_prior; & Create a coalescent constant-population tree prior \\
\verb;create_cep_tree_prior; & Create a coalescent exponential-population tree prior \\
\verb;create_yule_tree_prior; & Create a Yule tree prior \\
\hline
\verb;create_beta_distr; & Create a beta distribution \\
\verb;create_exp_distr; & Create an exponential distribution \\
\verb;create_gamma_distr; & Create a gamma distribution \\
\verb;create_inv_gamma_distr; & Create an inverse gamma distribution \\
\verb;create_laplace_distr; & Create a Laplace distribution \\
\verb;create_log_normal_distr; & Create a log-normal distribution \\
\verb;create_normal_distr; & Create a normal distribution \\
\verb;create_one_div_x_distr; & Create a 1/X distribution \\
\verb;create_poisson_distr; & Create a Poisson distribution \\
\verb;create_uniform_distr; & Create a uniform distribution \\
\hline
\end{tabular}
\caption{rbeast2's main functions}
\label{tab:functions}
\end{table}

@@

\verb;rbeast2; also uses the same default prior distributions as BEAUti 
for each of the site models, clock models and tree priors. 
For example, by default, a Yule tree prior assumes the birth rate 
follows a uniform distribution, 
from minus infinity to plus infinity. 
This assumption implies that negative and positive birth rates are just as likely, 
where a negative birth rate is biologically impossible (note that 
in practice, this usually works out just fine).
One may prefer an exponential distribution instead, 
as this would assume only positive birth rates, 
and makes high birth rates unlikely.

The following script shows how to do this in \verb;rbeast2;:

@@

Our initial motivation to create \verb;rbeast2; 
is that we wanted to fix the crown age of a phylogeny.
BEAUti assumes that a phylogeny has a crown age that needs to be jointly-estimated
with the phylogeny and other parameters. It does not allow for fixing
the crown age. Without \verb;rbeast2;, one needs to manually edit the BEAST2 
configuration file, which is tedious and prone to errors. 
Fixing the crown ages is especially useful for theoretical experiments,
as this allows for one less source of variation.

This is how to specify a fixed crown age with \verb;rbeast2;:

@@

\verb;rbeast2; allows for the same functionality of Tracer. 
Tracer, among others, shows the effective sample
sizes of each posterior's variables. As an MCMC run needs to
converge first, Tracer discards the first 10\% of all parameter
estimates. This is called the "burn-in". To remove the burn-in
in \verb;rbeast2;:

@@

Then, to calculate the effective sample sizes in \verb;rbeast2;:

@@

Tracer displays multiple summary statistics for each
estimated variable: the mean and its standard error, standard deviation,
variance, median, mode, geometric mean, 95\% highest posterior density interval, 
auto-correlation time and effective sample size. To obtain all these summary statistics 
of, for example, the estimated birth rate in \verb;rbeast2;:

@@

\verb;rbeast2; allows for the same functionality of DensiTree.
DensiTree displays the phylogenies in a posterior at the same
time scale over one another, allowing to see the uncertainty in
topology and branch lengths. To visualize the same in \verb;rbeast2;:

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{rbeast2 development and other resources}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\verb;rbeast2; is free, libre and open source software available from the official R package archive at 
\url{http://cran.r-project.org/src/contrib/PACKAGES.html\#rbeast2}
and is licensed under the GNU General Public License v3.0.

\verb;rbeast2; uses the Travis CI \cite{travis} 
continuous integration service, which is known to significantly 
increase the the number of bugs exposed \cite{vasilescu2015} and increases
the speed at which new features are added \cite{vasilescu2015}.
\verb;rbeast2; has a 100\% code coverage, which correlates with code quality \cite{horgan1994,del1995correlation}. 
\verb;rbeast2; follows Hadley Wickham's style guide \cite{style_guide}, which improves software quality \cite{fang2001}.
\verb;rbeast2; is dependent on multiple packages, which are 
\verb;APE; \cite{APE}, 
\verb;beautier; \cite{beautier},
\verb;beastier; \cite{beastier},
\verb;devtools; \cite{devtools},
\verb;geiger; \cite{GEIGER},
\verb;ggplot2; \cite{ggplot2},
\verb;knitr; \cite{knitr},
\verb;phangorn; \cite{phangorn},
\verb;rmarkdown; \cite{rmarkdown},
\verb;seqinr; \cite{seqinr},
\verb;stringr; \cite{stringr},
\verb;testit; \cite{testit} and 
\verb;tracerer; \cite{tracerer} and
\verb;TreeSim; \cite{TreeSim}.

\verb;rbeast2;'s development takes place on GitHub \cite{github},
\url{https://github.com/richelbilderbeek/rbeast2}, 
which accommodates collaboration \cite{perez2016ten} 
and improves transparency \cite{gorgolewski2016practical}.
\verb;rbeast2;'s GitHub facilitates feature requests and has guidelines how to do so.

\verb;rbeast2;'s documentation is extensive. All functions are documented
in the package's internal documentation. For quick use, 
each exported function shows a minimal example. 
For easy exploration, each exported function's documentation links to related functions.
Additionally, \verb;rbeast2; has a vignette that demonstrates extensively how
to use it. The GitHub documentation helps to get started, with a dozen examples 
of BEAUti screenshots with equivalent \verb;rbeast2; code.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Citation of rbeast2}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Scientists using \verb;rbeast2; in a published paper can cite this
article, and/or cite the \verb;rbeast2; package 
directly. To obtain this citation from within an R script, use:

\begin{lstlisting}[language=R]
> citation("rbeast2")
\end{lstlisting}
